Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system with plausible mechanism

### 2024-07-29 _tex/index.tex
Built site for gh-pages

\subsection{DAGs, SCMs, and the flow of association and
causation}\label{sec-background-dag}

>>>>

\subsection{DAGs, SCMs, and the flow of association and causation in
graphs}\label{sec-background-dag}

Graph theory is a branch of mathematics focused on the study of graphs.
Graphs are mathematical structures that model pairwise relations between
objects. They can represent physical relations, such as electrical
circuits and roadways, and less tangible structures, such as ecosystems
and sociological relations. Graphs have proven useful in various fields,
including computer science, operations research, and the natural and
social sciences \citep{Gross_et_al_2018}.

In statistics, one application that incorporates concepts from graph
theory is causal inference. Specifically, the structural approach to
causal inference uses DAGs to provide a formal and graphical
representation of a system's causal structure \citep{Neal_2020}. In this
context, a graph is a collection of nodes connected by edges, where
nodes represent random variables. In a directed graph, edges extend from
one node to another, with arrows indicating the direction of causal
influence. In a directed acyclic graph, the direction of causal
influences does not form a loop, ensuring that causal influences do not
cycle back on themselves \citep{McElreath_2020}.

Given the heuristic nature of DAGs, these are easier to understand using
a motivating example. Consider a system where variables \(X\) and \(Z\)
influence a third variable \(Y\). In this system, it is assumed that
\(X\) and \(Z\) are dependent on their own processes and are, therefore,
independent from each other. Figure~\ref{fig-dag1a} presents the
plausible causal structure of this system. The DAG shows the endogenous
variables \(V=\{X,Z,Y\}\) as circled black nodes, indicating these
variables are observed. The arrows connecting the variables indicate the
direction of causal influence, while a lack of influence is indicated by
the absence of arrows. Moreover, the exogenous variables
\(E=\{e_{X},e_{Z},e_{Y}\}\) represent everything else that is chosen not
to be modeled explicitly. These exogenous variables, or
\emph{disturbances}, are usually represented by open circles to indicate
their unobserved nature. Although this DAG explicitly shows the
exogenous variables, conventionally these are omitted for brevity,
resulting in an equivalent graph as shown in Figure~\ref{fig-dag1b}.

\begin{minipage}{0.50\linewidth}

\includegraphics[width=0.9\textwidth,height=\textheight]{images/figures/dag1a.png}

\subcaption{\label{fig-dag1a}Full DAG}

\end{minipage}%
%
\begin{minipage}{0.50\linewidth}

\includegraphics[width=0.35\textwidth,height=\textheight]{images/figures/dag1b.png}

\subcaption{\label{fig-dag1b}Simplified DAG}

\end{minipage}%

\caption{\label{fig-dags}DAGs for a plausible causal structure in a
system.}
